Fitting metabolic models to dissolved oxygen data: the Estuarine BAyesian Single-station Estimation method

### 2023-05-17 manu-draft.qmd
added JMA edits

Methods for estimating metabolism in aquatic environments are numerous and include open-water techniques, bottle-based incubations, ecoystem budgets, oxygen isotopes, and use of inert gases [@kemp2011; @staehr2011]. Based on the variety of assumptions and limitations of each method, comparisons of metabolic rates within and between ecosystems hav been challenging. For example, an assessment in Randers Fjord (Denmark) showed agreement in both the sign and magnitude among four different approaches [@gazeau2005]. However, gross primary production (P) and ecosystem respiration (R) estimates from incubations were consistently lower than those from open-water methods, consistent with the estuarine respiration synthesis of @hopkinson2005. Even within one technique, there are numerous choices to be made that alter the outcome. For example, in the commonly used ${}^{14}$C technique for measuring primary production, @cloern2014 noted that investigators have made divergent choices with regard to filtering mesozooplankton grazers, measuring ${}^{14}$C in dissolved organic carbon, accounting for changing spectral light quality with depth, and choosing the euphotic zone depth, the incubation time, and the incubation location (*in situ* vs. on-deck). These challenges are further compounded in estuaries where high temporal and spatial variability have been noted in global syntheses of P and R. @cloern2014 found that annual phytoplankton primary production can vary 10-fold within an estuary and up to 5-fold from year-to-year. @hopkinson2005 focused on cross-system variations and found benthic and pelagic respiration rates varying by more than factors of 30 and 40, respectively, ranges that are broadly consistent with whole system respiration rates based on the open-water method [@hoellein2013].

@@

As a final evaluation of EBASE to return known metabolic parameters, a new time series was created that added estimated effects of instrument and tidal noise to the synthetic time series. The objective of the analysis is to develop an expectation of how metabolic estimates may change using in situ data with noise that may substantially mask a biological signal in dissolved oxygen measurements. The instrument and tidal noise were estimated using a weighted regression analysis [@Beck15; @Beck21] on actual observed dissolved oxygen time series for Apalachicola Bay using data from 2021. This approach models dissolved oxygen as a function of time, salinity, and season using a moving window approach with weights of the independent variables appropriate for each time period within the windows. The results of this analysis return a predicted and detided dissolved oxygen time series. The instrument noise was estimated as the difference between the observed and predicted time series and the tidal noise was estimated as the difference between the predicted and detided time series. Both the instrument and tidal noise were added to the same synthetic time series above (@fig-synapanos). EBASE was then used with unconstrained priors (large $\sigma$) on the synthetic time series with noise, then compared with results from the original synthetic time series.

@fig-synapanoscmp shows the comparison of the EBASE results for the synthetic time series with instrument and tidal noise to those from the original synthetic time series. Results for both time series produced similar dissolved oxygen estimates (top-left panel, @fig-synapanoscmp), although notable differences were observed in the remaining parameters. In particular, gross primary production (P) and ecosystem respiration (R) were generally higher and more variable for results from the noisy time series, with differences in the former attributed to higher estimates in the $a$ parameter. As for the previous results, the $b$ parameter was estimated with low precision for both input time series, although results were centered around the true value of 0.251 $\left(\text{cm}/\text{hr}\right)/\left(\text{m}^2/\text{s}^2\right)$. Gas exchange (D) was similar between the input time series, suggesting lack of precision in the $b$ parameter regardless of noise had a minimal influence.

@@

@fig-apacmpfig and `r run_reference(id = "tbl-apacmptab")` provide a comparison of the metabolic estimates from the EBASE and Odum methods. Correlations for all metabolic estimates, regardless of the input time series, were positive and significantly correlated, as expected. However, correlations were generally stronger for the metabolic results based on the detided dissolved oxygen time series. The only exception for the correlation estimates was gas exchange (D), which had a slightly weaker correlation using the detided results. @fig-apacmpfig shows that the Odum results for gas exchange were more variable for the same values returned as no or small gas exchange for the EBASE method based on the detided dissolved oxygen. The RMSE values of each regression comparing the metabolic estimates between methods were generally lower in the detided results, excluding D which was higher with the detided dissolved oxygen relative to the observed.

@@

The second advantage of EBASE that makes the model more appropriate for estuarine settings included several modification to the underlying mass-balance equation for dissolved oxygen (@eq-ebase). As described in @sec-basemodif, EBASE is an estuarine adaptation of BASE (@eq-base) that was developed for stream applications and many assumptions of the underlying model of the latter were inappropriate for the former. Specifically, EBASE was modified to address three primary concerns of using BASE in an estuarine setting. First, the estimation of gross primary production as a function of PAR assumed a linear relationship such that the increase of production per unit increase of PAR was consistent for any level of the latter. The BASE model includes an exponential parameter, $p$, that defines a non-linear relationship between P and PAR (although it should be noted that BASE provides an option to set $p = 1$). @Holtgrieve10 describes a simple linear photosynthesis-irradiance curve as appropriate for systems that are never light-saturated and this assumption likely holds true for estuaries that are commonly more turbid than open-ocean or freshwater locations. Second, EBASE removed the temperature-dependency of respiration and assumed that respiration did not change with water temperature, consistent with applications of the Odum method to other shallow estuaries [@Beck15; @Murrell18]. This modification addresses recent work that suggests respiration in streams may be reduced during the night with lower temperatures as the respiration of labile autochthonous organic matter may contribute to changes in the dissolved oxygen time series [@sadro2014; @schindler2017]. There is currently no justification for this assumption in estuaries. Finally, the gas exchange in EBASE used the parameterization in @Wanninkhof14 that is primarily wind-based and developed for the open-ocean, which is clearly more appropriate than the prior formulation in BASE that includes a temperature and depth dependency for streams. Overall, these changes also reduced the required inputs and many monitoring locations nationwide include the requisite data for applying EBASE (i.e., dissolved oxygen, water temperature, salinity, PAR, and wind speed, as collected by the NERRS).

@@

While the EBASE theory and software provide advantages for metabolic estimates, the approach is not without limitations. Like most models, explicit choices are needed to apply EBASE to time series data. First, the use of unconstrained prior distributions was shown to reproduce known metabolic parameters with reasonable certainty, although the $b$ parameter was still relatively unstable and unconstrained priors may not provide useful estimates in estuarine settings outside of the location evaluated herein. This may require further constraining the priors, although doing so may produce biased results in the absence of additional data to justify a constraint. Field-based measurements or alternative metabolic models could be used for informing priors. As an example, initial application of BASE to estuarine data prior to developing EBASE demonstrated that unstable and unrealistic metabolic rates were returned with uninformed priors. The Odum method was used on the same data to develop an appropriate expectation of the range of values for the reaeration coefficient, $K$ , which was then used to inform the prior distribution using the BASE method. More stable and realistic metabolic estimates were then returned by BASE, demonstrating the value of the Bayesian approach that can incorporate prior knowledge. A similar approach could be applied to create more accurate estimates from EBASE depending on location. An additional choice for using EBASE is the model optimization period, although some guidance is provided above.

>>>>

Methods for estimating metabolism in aquatic environments are numerous and include open-water techniques, bottle-based incubations, ecoystem budgets, oxygen isotopes, and use of inert gases [@kemp2011; @staehr2011]. Based on the variety of assumptions and limitations of each method, comparisons of metabolic rates within and between ecosystems have been challenging. For example, an assessment in Randers Fjord (Denmark) showed agreement in both the sign and magnitude among four different approaches [@gazeau2005]. However, gross primary production (P) and ecosystem respiration (R) estimates from incubations were consistently lower than those from open-water methods, consistent with the estuarine respiration synthesis of @hopkinson2005. Even within one technique, there are numerous choices to be made that alter the outcome. For example, in the commonly used ${}^{14}$C technique for measuring primary production, @cloern2014 noted that investigators have made divergent choices with regard to filtering mesozooplankton grazers, measuring ${}^{14}$C in dissolved organic carbon, accounting for changing spectral light quality with depth, and choosing the euphotic zone depth, the incubation time, and the incubation location (*in situ* vs. on-deck). These challenges are further compounded in estuaries where high temporal and spatial variability have been noted in global syntheses of P and R. @cloern2014 found that annual phytoplankton primary production can vary 10-fold within an estuary and up to 5-fold from year-to-year. @hopkinson2005 focused on cross-system variations and found benthic and pelagic respiration rates varying by more than factors of 30 and 40, respectively, ranges that are broadly consistent with whole system respiration rates based on the open-water method [@hoellein2013].

@@

As a final evaluation of EBASE to return known metabolic parameters, a second synthetic time series was created that added estimated effects of instrument and tidal noise to the original synthetic time series. The objective of the analysis is to develop an expectation of how metabolic estimates may change using in situ data with noise that may substantially mask a biological signal in dissolved oxygen measurements. The instrument and tidal noise were estimated using a weighted regression analysis [@Beck15; @Beck21] on actual observed dissolved oxygen time series for Apalachicola Bay using data from 2021. This approach models dissolved oxygen as a function of time, salinity, and season using a moving window approach with weights of the independent variables appropriate for each time period within the windows. The results of this analysis return a predicted and detided dissolved oxygen time series. The instrument noise was estimated as the difference between the observed and predicted time series and the tidal noise was estimated as the difference between the predicted and detided time series. Both the instrument and tidal noise were added to the original synthetic time series above (@fig-synapanos). EBASE was then used with unconstrained priors (large $\sigma$) on the synthetic time series with noise, then compared with results from the original synthetic time series.

@fig-synapanoscmp shows the comparison of the EBASE results for the synthetic time series with instrument and tidal noise to those from the original synthetic time series. Results for both time series produced similar dissolved oxygen estimates (top-left panel, @fig-synapanoscmp), although notable differences were observed in the remaining parameters. In particular, gross primary production (P) and ecosystem respiration (R) were generally higher and more variable for results from the noisy time series, with differences in the former attributed to higher estimates in the $a$ parameter. As for the previous results, the $b$ parameter was estimated with low precision for both input time series, although results were centered around the true value of 0.251 $\left(\text{cm}/\text{hr}\right)/\left(\text{m}^2/\text{s}^2\right)$. Gas exchange (D) was similar between the input time series, suggesting that lack of precision in the $b$ parameter had a minimal influence regardless of noise.

@@

@fig-apacmpfig and `r run_reference(id = "tbl-apacmptab")` provide a comparison of the metabolic estimates from the EBASE and Odum methods. Correlations for all metabolic estimates, regardless of the input time series, were positive and significantly correlated, as expected. However, correlations were generally stronger for the metabolic results based on the detided dissolved oxygen time series. The only exception for the correlation estimates was gas exchange (D), which had a slightly weaker correlation using the detided results. @fig-apacmpfig shows that the Odum results for gas exchange were more variable than those for EBASE using the detided dissolved oxygen, where the latter often estimated gas exhange at much smaller values compared to the former. The RMSE values of each regression comparing the metabolic estimates between methods were generally lower in the detided results, excluding D which was higher with the detided dissolved oxygen relative to the observed.

@@

The second advantage of EBASE that makes the model more appropriate for estuarine settings included several modifications to the underlying mass-balance equation for dissolved oxygen (@eq-ebase). As described in @sec-basemodif, EBASE is an estuarine adaptation of BASE (@eq-base) that was developed for stream applications and many assumptions of the underlying model of the latter were inappropriate for the former. Specifically, EBASE was modified to address three primary concerns of using BASE in an estuarine setting. First, the estimation of gross primary production as a function of PAR assumed a linear relationship such that the increase of production per unit increase of PAR was consistent for any level of the latter. The BASE model includes an exponential parameter, $p$, that defines a non-linear relationship between P and PAR (although it should be noted that BASE provides an option to set $p = 1$). @Holtgrieve10 describes a simple linear photosynthesis-irradiance curve as appropriate for systems that are never light-saturated and this assumption likely holds true for estuaries that are commonly more turbid than open-ocean or freshwater locations. Second, EBASE removed the temperature-dependency of respiration and assumed that respiration did not change with water temperature, consistent with applications of the Odum method to other shallow estuaries [@Beck15; @Murrell18]. This modification addresses recent work that suggests respiration in streams may be reduced during the night with lower temperatures as the respiration of labile autochthonous organic matter may contribute to changes in the dissolved oxygen time series [@sadro2014; @schindler2017]. There is currently no justification for this assumption in estuaries. Finally, the gas exchange in EBASE used the parameterization in @Wanninkhof14 that is primarily wind-based and developed for the open-ocean, which is clearly more appropriate than the prior formulation in BASE that includes a temperature and depth dependency for streams. Overall, these changes also reduced the required inputs and many monitoring locations nationwide include the requisite data for applying EBASE (i.e., dissolved oxygen, water temperature, salinity, PAR, and wind speed, as collected by the NERRS).

@@

While the EBASE theory and software provide advantages for metabolic estimates, the approach is not without limitations. Like most models, explicit choices are needed to apply EBASE to time series data. First, the use of unconstrained prior distributions was shown to reproduce known metabolic parameters with reasonable certainty, although the $b$ parameter was still relatively unstable and unconstrained priors may not provide useful estimates in estuarine settings outside of the location evaluated herein. This may require further constraining the priors, although doing so may produce biased results in the absence of additional data to justify a constraint. Field-based measurements or alternative metabolic models could be used for informing priors. As an example, initial application of BASE to estuarine data prior to developing EBASE demonstrated that unstable and unrealistic metabolic rates were returned with uninformed priors. The Odum method was used on the same data to develop an appropriate expectation of the range of values for the reaeration coefficient, $K$, which was then used to inform the prior distribution using the BASE method. More stable and realistic metabolic estimates were then returned by BASE, demonstrating the value of the Bayesian approach that can incorporate prior knowledge. A similar approach could be applied to create more accurate estimates from EBASE depending on location. An additional choice for using EBASE is the model optimization period, although some guidance is provided above.
